---
title: "FAFSA submissions"
author: "Ian Hodgson"
date: "2024-03-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
rm(list = ls())
library(tidyverse)
library(readxl)
```

``` {r clean data, include = F}

# IMPORT DATA ----

### Grade 12 enrollment
gr12_2223 <- read_xlsx("data/FAFSA/2223MembBySchoolByGrade.xlsx",
                       sheet = "School", 
                       skip = 2,
                       na = "*") %>% 
  janitor::clean_names() %>% 
  select(1:4, x12) %>% 
  rename(students = x12) %>% 
  filter(!is.na(students)) %>% 
  mutate(class = "2023")

gr12_2324 <- read_xlsx("data/FAFSA/2324MembBySchoolByGrade.xlsx",
                       sheet = "School", 
                       skip = 2,
                       na = "*") %>% 
  janitor::clean_names() %>% 
  select(1:4, x12) %>% 
  rename(students = x12) %>% 
  filter(!is.na(students)) %>% 
  mutate(class = "2024")

gr12_enroll <- bind_rows(gr12_2223, 
                         gr12_2324)

## state to NCES code crosswalk 
nces_xwalk <- read_xlsx("data/FAFSA/ncesdata_441DE4E0.xlsx",
                       skip = 11,
                       na = "†") %>% 
  janitor::clean_names() %>% 
  mutate(school_number = str_extract(state_school_id, "[0-9]{4}$") %>% 
           as.numeric,
         district_number = str_extract(state_district_id, "[0-9]{2}$") %>% 
           as.numeric)

## FAFSA submissions
fafsa <- read_xls("data/FAFSA/2024 reports/FL-FAFSA submissions-2024-03-08.xls",
                  sheet = "FL School Level Data",
                  skip = 4, 
                  col_names = F, 
                  na = "<5") %>% 
  mutate(date = ymd("2024-03-08")) %>% 
  bind_rows(read_xls("data/FAFSA/2024 reports/FL-FAFSA submissions-2024-03-15.xls",
                  sheet = "FL School Level Data",
                  skip = 4, 
                  col_names = F, 
                  na = "<5") %>% 
              mutate(date = ymd("2024-03-15")))

names(fafsa) <- c("school_code",
                  "school_name", 
                  "city", 
                  "state", 
                  "submitted_2024",
                  "completed_2024",
                  "submitted_2023",
                  "completed_2023", 
                  "submitted_2023_june", 
                  "completed_2023_june",
                  "submitted_2023_dec",
                  "completed_2023_dec",
                  "date")

# COMBINE DATA ----

fafsa_subs <- fafsa %>% 
  select(1:6,13) %>% 
  rename(submitted = submitted_2024,
         completed = completed_2024) %>% 
  mutate(class = "2024") %>%
  bind_rows(fafsa %>% 
              select(1:4, 7:8, 13) %>% 
              rename(submitted = submitted_2023,
                     completed = completed_2023) %>% 
              mutate(class = "2023")) %>% 
  mutate(submitted = as.numeric(submitted), 
         completed = as.numeric(completed)) %>% 
  group_by(school_code, school_name, state, class, date) %>% 
  summarise(submitted = sum(submitted, na.rm = T),
            completed = sum(completed, na.rm = T)) %>% 
  ungroup() %>% 
  left_join(nces_xwalk %>% 
              select(school_code = nces_school_id,
                     school_number, 
                     district_number,
                     type, 
                     charter),
            by = "school_code") %>% 
  left_join(gr12_enroll, 
            by = c("school_number", 
                   "district_number", 
                   "class")) %>% 
  filter(!is.na(type),
         charter == "No",
         !is.na(students)) %>% 
  mutate(submission_rate = submitted/students)

```

``` {r district submission rates, include = T}
district <- fafsa_subs %>% 
  filter(date == max(date)) %>% 
  group_by(class, district) %>% 
  summarise(submitted = sum(submitted, na.rm = T),
            students = sum(students)) %>% 
  mutate(submit_share = submitted/students) %>% 
  ungroup() %>% 
  group_by(district) %>% 
  arrange(district, class) %>% 
  mutate(change = (submit_share/submit_share[1])-1)

district %>% filter(class == "2024")

state <- fafsa_subs %>% 
  group_by(class) %>% 
  summarise(submitted = sum(submitted, na.rm = T),
            students = sum(students)) %>% 
  mutate(submit_share = submitted/students) %>% 
  mutate(change = (submit_share/submit_share[1])-1)

state
  

```

### Completion rates over time

``` {r clean archive data, include = F}
## Cycle 2024-25
cycle_2425 <- read_xlsx("data/FAFSA/archived reports/HSARCHIVE_combined.xlsx",
                        sheet = "cycle 2425", 
                        na = "<5") %>% 
  janitor::clean_names() %>% 
  pivot_longer(cols = 5:6, 
               values_to = "submitted",
               names_to = "date") %>% 
  mutate(class = "2024",
         month = mdy(date) %>% month(),
         date = date %>% mdy()) %>% 
    filter(state == "FL") %>% 
  filter(!is.na(submitted))

## Cycle 2023-24
cycle_2324 <- read_xlsx("data/FAFSA/archived reports/HSARCHIVE_combined.xlsx",
                        sheet = "cycle 2324", 
                        na = "<5") %>% 
  janitor::clean_names() %>% 
  pivot_longer(cols = 5:12, 
               values_to = "submitted",
               names_to = "date") %>% 
  mutate(class = "2023",
         month = date %>% mdy() %>% month(),
         date = date %>% mdy) %>% 
    filter(state == "FL") %>% 
  filter(!is.na(submitted))

temp <- cycle_2324 %>% 
  bind_rows(cycle_2425) %>% 
  left_join(fafsa_subs %>% distinct(school_code, class, students, district),
            by = c("school_code", "class")) %>% 
  mutate(submission_rate = as.numeric(submitted)/students,
         submitted = as.numeric(submitted)) %>% 
  rename(school_name = name) %>% 
  bind_rows(fafsa_subs %>% 
              mutate(date = ifelse(class == "2023", 
                                   date - years(1),
                                   date) %>% 
                       as.Date())) %>%
  select(school_code, school_name, date, submitted, class, month, students,
         submission_rate, district, date) %>% 
  arrange(school_name, date)

temp_dist <- temp %>% 
  group_by(class, month, district, date) %>% 
  filter(!is.na(students)) %>% 
  summarise(submitted = sum(submitted, na.rm = T),
            students = sum(students, na.rm = T)) %>% 
  mutate(submission_rate = submitted/students,
         date_adj = ifelse(class == "2023",
                           date, 
                           date - years(1)) %>% 
           as.Date(),
         date_adj = ifelse(is.na(date_adj),
                           ymd("2023-02-28"),
                           date_adj)) %>% 
  ungroup()

df_out_district <- temp_dist %>% 
  filter(class == "2023") %>% 
  select(district, date_adj,
         submitted_2023 = submitted, 
         students_2023 = students, 
         submission_rate_2023 = submission_rate) %>% 
  left_join(temp_dist %>% 
              filter(class == "2024") %>% 
              select(district, date_adj,
                     submitted_2024 = submitted, 
                     students_2024 = students, 
                     submission_rate_2024 = submission_rate),
            by = c("district", "date_adj")) %>% 
  mutate(date_adj = date_adj %>% as.Date) %>% 
  arrange(date_adj, district)

df_out_state <- df_out_district %>% 
  group_by(date_adj) %>% 
  select(date_adj, 
         starts_with("students"), 
         starts_with("submitted")) %>% 
  summarise_all(.funs = function(x){sum(x, na.rm = T)}) %>% 
  mutate(submission_rate_2023 = submitted_2023/students_2023, 
         submission_rate_2024 = submitted_2024/students_2024) %>% 
  write_csv("data/FAFSA/output/submission_rate_state.csv", na = "")

df_out_district %>% 
  bind_rows(tibble(district = unique(df_out_district$district), 
                   date_adj =  rep(ymd("2022-09-30"), 70), 
                   submission_rate_2023 = rep(0, 70))) %>% 
  mutate(submission_rate_2024 = ifelse(date_adj == ymd("2022-12-31"),
                                       0,
                                       submission_rate_2024)) %>% 
  filter(district %in% c("HILLSBOROUGH", 
                         "PINELLAS",
                         "PASCO")) %>% 
  ggplot() + 
  geom_line(aes(x = date_adj, 
                y = submission_rate_2023, 
                title = "2023"), 
            col = "red") + 
  geom_line(aes(x = date_adj, 
                y = submission_rate_2024, 
                title = "2024"),
            col = "blue") + 
  theme_minimal() +
  facet_wrap(facets = "district", 
             ncol = 2)
  

```



### Submission rates by district, broken down by poverty and race

``` {r pov and race, include = F}
lunch_status <- read_xlsx("data/FAFSA/2324FS2-Lunch-Status-School.xlsx",
                          sheet = 1, skip = 2, na = "*") %>% 
  janitor::clean_names() %>% 
  group_by(district) %>% 
  mutate(dist_ave_rate = mean(rate_wo_multiplier),
         rate_g_ave = rate_wo_multiplier > dist_ave_rate,
         rate_g_X = rate_wo_multiplier >= .5,
         pov_quartile = case_when(
           rate_wo_multiplier <= .3463 ~ "bottom",
           rate_wo_multiplier >= .5848 ~ "top",
           !is.na(rate_wo_multiplier) ~ "middle",
           T ~ "ERROR")) %>% 
  mutate(district_number = as.numeric(dist_number),
         school_number = as.numeric(schl_number)) %>% 
    select(district_number, school_number, school_name, 
           lunch_rate_wo_multiplier = rate_wo_multiplier, 
           lunch_rate_g_ave = rate_g_ave, 
           lunch_rate_g_X = rate_g_X,
           lunch_quartile = pov_quartile)

subs_dis_pov <- fafsa_subs %>% 
  filter(class == "2024",
         !is.na(submission_rate)) %>% 
  left_join(lunch_status) 

subs_dis_pov %>% 
  ungroup() %>% 
  group_by(lunch_quartile) %>% 
  summarise(submission_rate = mean(submission_rate),
            count = n())

```

There appears to be a strong relationship between family income and FAFSA submission rates. The schools with the highest share of low-income students (top quartile, share >= 58.5%) had 38% lower submission rate than the most afflent schools (bottom quartile, share <= 34.6%)

``` {r poverty regression, include = T}
subs_dis_pov %>% 
  # filter(district %in% c("HILLSBOROUGH",
  #                        "PINELLAS",
  #                        "PASCO")) %>% 
  ggplot() + 
  geom_point(aes(x = lunch_rate_wo_multiplier, 
                 y = submission_rate, size = students), 
             alpha = .2) + 
  geom_smooth(aes(x = lunch_rate_wo_multiplier, 
                 y = submission_rate),
              method = "lm", 
              formula = y~x) + 
  theme_minimal()
```            
``` {r submissions by rate, include = F}
school_race <- read_xlsx("data/FAFSA/2324MembBySchoolByGradeByRace.xlsx",
                          sheet = 3, skip = 2, na = "*") %>% 
  janitor::clean_names() %>% 
  filter(grade == "12")

subs_sch_race <- left_join(fafsa_subs %>% filter(class == "2024"),
                           school_race) %>% 
  filter(!is.na())

```
